---
title: "Tutorial for synergy prediction using the DrugLogics software pipeline"
author: "[John Zobolas](https://github.com/bblodfon)"
date: "Last updated: `r format(Sys.time(), '%d %B, %Y')`"
description: "A software tutorial"
url: 'https\://druglogics.github.io/synergy-tutorial/'
github-repo: "druglogics/synergy-tutorial"
bibliography: references.bib
link-citations: true
site: bookdown::bookdown_site
---

# Intro {-}

The purpose of the present tutorial is to provide guidance for the installation, execution, visualization and interpretation of output simulation results for the DrugLogics software pipeline module [druglogics-synergy](https://github.com/druglogics/druglogics-synergy).

The `druglogics-synergy` module runs sequentially two software modules.
Firstly, [gitsbe](https://github.com/druglogics/gitsbe) is used to create an ensemble of Boolean models fit to a specific steady state or perturbation data using a genetic parameterization algorithm.
Then [drabme](https://github.com/druglogics/drabme) uses the output models from gitsbe to perform a drug response analysis of a given drug panel and produces predicted synergy scores for each respective drug combination.

You can find the full documentation for these two modules in the following link: https://druglogics.github.io/druglogics-doc/.

# Install {-}

:::{.green-box}
Firstly, make sure you have installed [Maven](https://maven.apache.org/install.html) `3.6.0` and `Java 8` (minimum working versions).
For the rest of the tutorial we are going to use the `druglogics-synergy` at version `1.2.1`, which includes `gitsbe` at `1.3.1` version and `drabme` at `1.2.1` version.
:::

There are **two ways** to install `druglogics-synergy`.

1. **The easiest option** is to download the latest released package from GitHub: https://github.com/druglogics/druglogics-synergy/packages/.
The file of interest is `synergy-1.2.1-jar-with-dependencies.jar`, which includes all dependencies and does not require any manual installation whatsoever.

2. **The other option** is to clone the repositories ^[Due to GitHub's security policies, not allowing users to download a public package without a PAT (Personal Access Token) as of November 2021, you have to install each dependency repository (gitsbe, drabme) separately.] and install them manually using Maven:

```
git clone https://github.com/druglogics/gitsbe.git
cd gitsbe
git checkout v1.3.1
mvn clean install
cd ../

git clone https://github.com/druglogics/drabme.git
cd drabme
git checkout v1.2.1
mvn clean install
cd ../

git clone https://github.com/druglogics/druglogics-synergy
cd druglogics-synergy
git checkout v1.2.1
mvn clean install
```

After executing the above commands, the `synergy-1.2.1-jar-with-dependencies.jar` file will be created inside the `druglogics-synergy/target` directory.

:::{.note}
Even if you choose the 1st option, make sure you clone the `druglogics-synergy` repo and put the `synergy-1.2.1-jar-with-dependencies.jar` file inside the `druglogics-synergy/target` directory (as `mvn clean install` command does) to be inline with the next instructions in this tutorial.
:::

# Run CASCADE 1.0 Example {-}

We are going to use the [`ags_cascade_1.0`](https://github.com/druglogics/druglogics-synergy/tree/v1.2.1/ags_cascade_1.0) directory that resides in the `druglogics-synergy` directory as an input to the `Launcher` class to run the CASCADE 1.0 example.

The command to execute from the root of the `druglogics-synergy` directory is the following:

```
java -cp target/synergy-1.2.1-jar-with-dependencies.jar eu.druglogics.synergy.Launcher --inputDir=ags_cascade_1.0
```

The above command should create a new directory inside `ags_cascade_1.0` that includes the output result files.
Given that the simulations use all available cores, it shouldn't take more than **30sec to 1min** to complete the execution on a modern computer.

:::{.note}
For the rest of this tutorial we are going to use the output results produced from the execution of the above command, please refer to [this directory](https://github.com/druglogics/synergy-tutorial/tree/main/ags_cascade_1.0_20211121_150114).
:::

## Inputs {-}

A brief description of the input files included in the [`ags_cascade_1.0`](https://github.com/druglogics/druglogics-synergy/tree/v1.2.1/ags_cascade_1.0) directory follows:

1. [`network.sif`](https://github.com/druglogics/druglogics-synergy/tree/v1.2.1/ags_cascade_1.0/network.sif): a single-interactions network file in Cytoscape's .sif tab-delimited format.
This file defines the CASCADE 1.0 topology.
2. [`training`](https://github.com/druglogics/druglogics-synergy/tree/v1.2.1/ags_cascade_1.0/training): this file has the training data for gitsbe's algorithm in one of the formats specified in the respective [documentation](https://druglogics.github.io/druglogics-doc/training-data.html).
Here, **we train to an unperturbed condition, with a steady state observation response** as curated from many publications for the AGS cell line [@Flobak2015].
3. [`modeloutputs`](https://github.com/druglogics/druglogics-synergy/tree/v1.2.1/ags_cascade_1.0/modeloutputs): this is a file that is used to calculate the output growth response of a Boolean model after its attractors are computed.
See more information in the respective [documentation](https://druglogics.github.io/druglogics-doc/modeloutputs.html).
Here, we define 3 nodes that signal cell proliferation when active (`RSK_f`, `MYC` and `TCF7_f`) and 3 that signal apoptosis (`CASP8`, `CASP9` and `FOXO_f`).
4. [`drugpanel`](https://github.com/druglogics/druglogics-synergy/tree/v1.2.1/ags_cascade_1.0/drugpanel): this file has the drugs and their respective targets, which are going to be analyzed by drabme using gitsbe's output Boolean models.
More info can be found [here](https://druglogics.github.io/druglogics-doc/drug-panel.html).
5. [`config`](https://github.com/druglogics/druglogics-synergy/tree/v1.2.1/ags_cascade_1.0/config): configuration file that defines several parameters used by gitsbe and drabme.
Apart from the short description for each parameter provided in the `config` file, there is a more complete documentation of each option [available here](https://druglogics.github.io/druglogics-doc/gitsbe-config.html).
    - The most important options for [gitsbe](https://github.com/druglogics/druglogics-synergy/blob/v1.2.1/ags_cascade_1.0/config#L27) are the tool used to calculate the models' attractors (bioLQM, see [@Naldi2018]), **the number of simulations** $(50)$, number of generations per simulation ($20$), number of models per generation $(20)$, **number of best-fit models to save** after each simulation is finished $(3)$ and the type of [model mutations](https://druglogics.github.io/druglogics-doc/gitsbe-config.html#mutation-types) used (here we use *balance* or *link operator* mutations).
    - For [drabme](https://github.com/druglogics/druglogics-synergy/blob/v1.2.1/ags_cascade_1.0/config#L96), the two most important configuration parameters are the maximum drug set size to test ($2$, i.e. up to drug pairs) and the method used for the calculation of drug synergies (we use HSA [@gaddum1940pharmacology] in the `config` file, but also Bliss [@Bliss1939] is available).

## Outputs {-}

The most significant outputs in the [results directory](https://github.com/druglogics/synergy-tutorial/tree/main/ags_cascade_1.0_20211121_150114) are:

- **Gitsbe-specific**:
    - A [`models`](https://github.com/druglogics/synergy-tutorial/tree/main/ags_cascade_1.0_20211121_150114/models) directory with files in `.gitsbe` format ^[You can configure gitsbe to export the resulting Boolean models to more standard formats, see the respective [option here](https://github.com/druglogics/druglogics-synergy/blob/v1.2.1/ags_cascade_1.0/config#L88)], representing the best-fit boolean models out of the simulations of the genetic algorithm.
    Here, since we configured to have $50$ simulations and save the best $3$ models out of each simulation, we end up with a total of $150$ models after gitsbe finishes its execution.
- **Drabme-specific**:
  - A [`ensemble-wise synergies file`](https://github.com/druglogics/synergy-tutorial/blob/main/ags_cascade_1.0_20211121_150114/ags_cascade_1.0_ensemblewise_synergies.tab), providing the synergy predictions based on gitsbe's models.
  For every drug combination tested (we have $7$ drugs in the [drugpanel file](https://github.com/druglogics/druglogics-synergy/blob/v1.2.1/ags_cascade_1.0/drugpanel), so a total of ${7\choose2} = 21$ drug pairs are tested), drabme calculates a synergy score indicating how synergistic that combination was (more negative means more synergistic), based on the HSA synergy metric.

## How are predictions calculated? {-}

Let's look at an example: drabme [has calculated](https://github.com/druglogics/synergy-tutorial/blob/main/ags_cascade_1.0_20211121_150114/ags_cascade_1.0_ensemblewise_synergies.tab#L11) a synergy score equal to $\approx -0.943$ for the combination `PD-AK` (MEK and AKT inhibitor).

Drabme used gitsbe's $150$ output Boolean models and perturbed them first with the single drug `PD` (changing the equation of the respective target to equal $0$, i.e. `MEK_f = 0`).
Then it calculated all the models' attractors.
The ensemble of models that had at least one attractor were able to provide an [output response](https://druglogics.github.io/druglogics-doc/modeloutputs.html) each, from which drabme calculates the ensemble's average response to the `PD` drug.
Same procedure is followed for the `AK` drug: changing its target's equation appropriately (`AKT_f = 0`) in all $150$ models, calculating their attractors and using only the models that had at least one attractor to calculate the ensemble's average response to drug `PD`.
So we are left with two average single-drug response values, from which we choose as a *reference value* (according to the HSA model we used) **the more negative of the two** (as the values represent cell growth output).

Drabme repeats the previous process by now **changing both equations** in all $150$ models, computing their attractors and finding the ensemble's average growth output response for the `PD-AK` combination.
If this combination value is *lower than the reference value*, the difference is outputted in the ensemble synergies output file.
For our particular case, the difference of the combination's `PK-AD` average response to the minimum of the single-drug average responses in the $150$ model ensemble, is equal to $\approx -0.943$.
Since this value is the lowest in the output [ensemble-wise synergies file](https://github.com/druglogics/synergy-tutorial/blob/main/ags_cascade_1.0_20211121_150114/ags_cascade_1.0_ensemblewise_synergies.tab), `PK-AD` is the most synergistic drug combination out of our simulations.

# Visualize Prediction Performance {-}

In this section we assess the quality of drabme's synergy predictions by plotting ROC and PR curves.

## Observed synergies for CASCADE 1.0 {-}

The synergy prediction problem is a standard binary classification problem (synergy or no synergy, varying a threshold defining how stringent is our definition of synergy).
To properly assess our algorithm's performance we need to compare our results with a true set of drug synergies.
All $21$ drug combinations from the [drugpanel file](https://github.com/druglogics/druglogics-synergy/blob/v1.2.1/ags_cascade_1.0/drugpanel) were experimentally tested in [@Flobak2015], producing a set of $4$ gold standard synergies (see [respective file](https://github.com/druglogics/synergy-tutorial/blob/main/observed_synergies_cascade_1.0)): `PI-PD`, `PI-5Z`, `PD-AK` and `AK-5Z`.

## Workflow {-}

:::{.note}
We will use some `R` code to parse drabme's output prediction scores and visualize our pipeline's synergy prediction performance using the **Receiver Operating Characteristic (ROC)** and the **Precision-Recall (PR)** curves as well as the **Area Under the Curve (AUC)**.
The below code should be run from the root of the [`synergy-tutorial`](https://github.com/druglogics/synergy-tutorial) repository.

We also include a separate [R script](https://github.com/druglogics/synergy-tutorial/blob/main/code.R) with all the code in the below code blocks for ease of use.
:::

## Read Output Results {-}

Load the appropriate libraries:
```{r load-lib, class.source = 'fold-show'}
suppressMessages(library(dplyr))
library(tibble)
library(emba)
library(usefun)
library(PRROC)
library(DT)
```

Read the data using the `emba` library [@Zobolas2020]:
```{r read-data, class.source = 'fold-show'}
# Read ensemble-wise synergies file
# `ss` => models trained to steady state
ss_hsa_file = "ags_cascade_1.0_20211121_150114/ags_cascade_1.0_ensemblewise_synergies.tab"
ss_hsa_ensemblewise_synergies = emba::get_synergy_scores(ss_hsa_file)

# Read observed synergies file
observed_synergies_file = 'observed_synergies_cascade_1.0'
observed_synergies = emba::get_observed_synergies(observed_synergies_file)
# 1 (positive/observed synergy) or 0 (negative/not observed) for all tested drug combinations
observed = sapply(ss_hsa_ensemblewise_synergies$perturbation %in% observed_synergies, as.integer)

# Make a data table
pred_hsa = dplyr::bind_cols(ss_hsa_ensemblewise_synergies %>% rename(ss_score = score),
  tibble::as_tibble_col(observed, column_name = "observed"))
```

:::{#ss-pred-tbl}
Visualize the prediction results in a table format:
:::
```{r ss-pred-res}
DT::datatable(data = pred_hsa, options = 
  list(pageLength = 7, lengthMenu = c(7, 14, 21), searching = FALSE,
    order = list(list(2, 'asc')))) %>%
  DT::formatRound(columns = 2, digits = 5) %>%
  DT::formatStyle(columns = 'observed',
    backgroundColor = styleEqual(c(0, 1), c('white', 'yellow')))
```

## Receiver Operating Characteristic (ROC) {-}

We use the `get_roc_stats()` function from the `usefun` R package [@R-usefun]:
```{r roc-res, fig.align='center', out.width='65%', fig.width=5, fig.height=5, dpi=300, fig.cap='ROC curve (CASCADE 1.0, HSA synergy method, 150 models calibrated to the AGS steady state)', class.source = 'fold-show'}
# Get ROC statistics (`res_ss_ew$AUC` holds the ROC AUC)
res_ss_ew = usefun::get_roc_stats(df = pred_hsa, pred_col = "ss_score", label_col = "observed")

# Plot ROC
my_palette = RColorBrewer::brewer.pal(n = 9, name = "Set1")

plot(x = res_ss_ew$roc_stats$FPR, y = res_ss_ew$roc_stats$TPR,
  type = 'l', lwd = 3, col = my_palette[1], main = 'ROC curve, Ensemble-wise synergies (HSA)',
  xlab = 'False Positive Rate (FPR)', ylab = 'True Positive Rate (TPR)')
legend('bottomright', title = 'AUC', col = my_palette[1:2], pch = 19,
  legend = paste(round(res_ss_ew$AUC, digits = 2), "Calibrated"), cex = 1.3)
grid(lwd = 0.5)
abline(a = 0, b = 1, col = 'lightgrey', lty = 'dotdash', lwd = 1.2)
```

## Precision-Recall (PR) {-}

We use the `pr.curve()` function from the `PRROC` package [@Grau2015]:
```{r pr-res, fig.align='center', out.width='65%', fig.width=5, fig.height=5, dpi=300, fig.cap='PR curve (CASCADE 1.0, HSA synergy method, 150 models calibrated to the AGS steady state)', class.source = 'fold-show'}
# NOTE: PRROC considers by default that larger prediction values indicate the 
# positive class labeling. For us, the synergy scores belonging to the positive 
# or synergy class (observed = 1) are the lower ones, so we need to
# reverse the scores to correctly calculate the PR curve
pr_ss_hsa = PRROC::pr.curve(scores.class0 = pred_hsa %>% pull(ss_score) %>% (function(x) {-x}), 
  weights.class0 = pred_hsa %>% pull(observed), curve = TRUE, rand.compute = TRUE)

plot(pr_ss_hsa, main = 'PR curve, Ensemble-wise synergies (HSA)', 
  auc.main = FALSE, color = my_palette[1], rand.plot = TRUE)
legend('topright', title = 'AUC', col = my_palette[1:2], pch = 19,
  legend = paste(round(pr_ss_hsa$auc.davis.goadrich, digits = 2), "Calibrated"))
grid(lwd = 0.5)
```

# Normalization {-}

# R session info {-}

```{r session info, comment="", class.source = 'fold-show'}
xfun::session_info()
```

# References {-}
