---
title: "Tutorial for synergy prediction using the DrugLogics software pipeline"
author: "[John Zobolas](https://github.com/bblodfon)"
date: "Last updated: `r format(Sys.time(), '%d %B, %Y')`"
description: "A software tutorial"
url: 'https\://druglogics.github.io/synergy-tutorial/'
github-repo: "druglogics/synergy-tutorial"
bibliography: references.bib
link-citations: true
site: bookdown::bookdown_site
---

# Intro {-}

The purpose of the present tutorial is to provide guidance for the installation, execution, visualization and interpretation of output simulation results for the DrugLogics software pipeline module [druglogics-synergy](https://github.com/druglogics/druglogics-synergy).

The `druglogics-synergy` module runs sequentially two software modules.
Firstly, [gitsbe](https://github.com/druglogics/gitsbe) is used to create an ensemble of Boolean models fit to a specific steady state or perturbation data using a genetic parameterization algorithm.
Then [drabme](https://github.com/druglogics/drabme) uses the output models from gitsbe to perform a drug response analysis of a given drug panel and produces predicted synergy scores for each respective drug combination.

You can find the full documentation for these two modules in the following link: https://druglogics.github.io/druglogics-doc/.

# Install {-}

:::{.green-box}
Firstly, make sure you have installed [Maven](https://maven.apache.org/install.html) `3.6.0` and `Java 8` (minimum working versions).
For the rest of the tutorial we are going to use the `druglogics-synergy` at version `1.2.1`, which includes `gitsbe` at `1.3.1` version and `drabme` at `1.2.1` version.
:::

There are **two ways** to install `druglogics-synergy`.

1. **The easiest option** is to download the latest released package from GitHub: https://github.com/druglogics/druglogics-synergy/packages/.
The file of interest is `synergy-1.2.1-jar-with-dependencies.jar`, which includes all dependencies and does not require any manual installation whatsoever.

2. **The other option** is to clone the repositories ^[Due to GitHub's security policies, not allowing users to download a public package without a PAT (Personal Access Token) as of November 2021, you have to install each dependency repository (gitsbe, drabme) separately.] and install them manually using Maven:

```
git clone https://github.com/druglogics/gitsbe.git
cd gitsbe
git checkout v1.3.1
mvn clean install
cd ../

git clone https://github.com/druglogics/drabme.git
cd drabme
git checkout v1.2.1
mvn clean install
cd ../

git clone https://github.com/druglogics/druglogics-synergy
cd druglogics-synergy
git checkout v1.2.1
mvn clean install
```

After executing the above commands, the `synergy-1.2.1-jar-with-dependencies.jar` file will be created inside the `druglogics-synergy/target` directory.

:::{.note}
Even if you choose the 1st option, make sure you clone the `druglogics-synergy` repo and put the `synergy-1.2.1-jar-with-dependencies.jar` file inside the `druglogics-synergy/target` directory (as `mvn clean install` command does) to be inline with the next instructions in this tutorial.
:::

# Run CASCADE 1.0 Example {-}

We are going to use the [`ags_cascade_1.0`](https://github.com/druglogics/druglogics-synergy/tree/v1.2.1/ags_cascade_1.0) directory that resides in the `druglogics-synergy` directory as an input to the `Launcher` class to run the CASCADE 1.0 example.
Note that preparing an input directory with all necessary files is the easiest way to run `druglogics-synergy`.

The command to execute from the root of the `druglogics-synergy` directory is the following:

```
java -cp target/synergy-1.2.1-jar-with-dependencies.jar eu.druglogics.synergy.Launcher --inputDir=ags_cascade_1.0
```

The above command should create a new directory inside `ags_cascade_1.0` that includes the output result files.
Given that the simulations use all available cores, it shouldn't take more than **30sec to 1min** to complete the execution on a modern computer.

:::{.note}
For the rest of this tutorial we are going to use the output results produced from the execution of the above command, please refer to [this directory](https://github.com/druglogics/synergy-tutorial/tree/main/ags_cascade_1.0_20211121_150114).
:::

## Inputs {-}

A brief description of the input files included in the [`ags_cascade_1.0`](https://github.com/druglogics/druglogics-synergy/tree/v1.2.1/ags_cascade_1.0) directory follows:

1. [`network.sif`](https://github.com/druglogics/druglogics-synergy/tree/v1.2.1/ags_cascade_1.0/network.sif): a single-interactions network file in Cytoscape's .sif tab-delimited format.
This file defines the CASCADE 1.0 topology.
2. [`training`](https://github.com/druglogics/druglogics-synergy/tree/v1.2.1/ags_cascade_1.0/training): this file has the training data for gitsbe's algorithm in one of the formats specified in the respective [documentation](https://druglogics.github.io/druglogics-doc/training-data.html).
Here, **we train to an unperturbed condition, with a steady state observation response** as curated from many publications for the AGS cell line [@Flobak2015].
3. [`modeloutputs`](https://github.com/druglogics/druglogics-synergy/tree/v1.2.1/ags_cascade_1.0/modeloutputs): this is a file that is used to calculate the output growth response of a Boolean model after its attractors are computed.
See more information in the respective [documentation](https://druglogics.github.io/druglogics-doc/modeloutputs.html).
Here, we define 3 nodes that signal cell proliferation when active (`RSK_f`, `MYC` and `TCF7_f`) and 3 that signal apoptosis (`CASP8`, `CASP9` and `FOXO_f`).
4. [`drugpanel`](https://github.com/druglogics/druglogics-synergy/tree/v1.2.1/ags_cascade_1.0/drugpanel): this file has the drugs and their respective targets, which are going to be analyzed by drabme using gitsbe's output Boolean models.
More info can be found [here](https://druglogics.github.io/druglogics-doc/drug-panel.html).
5. [`config`](https://github.com/druglogics/druglogics-synergy/tree/v1.2.1/ags_cascade_1.0/config): configuration file that defines several parameters used by gitsbe and drabme.
Apart from the short description for each parameter provided in the `config` file, there is a more complete documentation of each option [available here](https://druglogics.github.io/druglogics-doc/gitsbe-config.html).
    - The most important options for [gitsbe](https://github.com/druglogics/druglogics-synergy/blob/v1.2.1/ags_cascade_1.0/config#L27) are the tool used to calculate the models' attractors (bioLQM, see [@Naldi2018]), **the number of simulations** $(50)$, number of generations per simulation ($20$), number of models per generation $(20)$, **number of best-fit models to save** after each simulation is finished $(3)$ and the type of [model mutations](https://druglogics.github.io/druglogics-doc/gitsbe-config.html#mutation-types) used (here we use *balance* or *link operator* mutations).
    - For [drabme](https://github.com/druglogics/druglogics-synergy/blob/v1.2.1/ags_cascade_1.0/config#L96), the two most important configuration parameters are the maximum drug set size to test ($2$, i.e. up to drug pairs) and the method used for the calculation of drug synergies (we use HSA [@gaddum1940pharmacology] in the `config` file, but also Bliss [@Bliss1939] is available).

## Outputs {-}

The most significant outputs in the [results directory](https://github.com/druglogics/synergy-tutorial/tree/main/ags_cascade_1.0_20211121_150114) are:

- **Gitsbe-specific**:
    - A [`models`](https://github.com/druglogics/synergy-tutorial/tree/main/ags_cascade_1.0_20211121_150114/models) directory with files in `.gitsbe` format ^[You can configure gitsbe to export the resulting Boolean models to more standard formats, see the respective [option here](https://github.com/druglogics/druglogics-synergy/blob/v1.2.1/ags_cascade_1.0/config#L88)], representing the best-fit boolean models out of the simulations of the genetic algorithm.
    Here, since we configured to have $50$ simulations and save the best $3$ models out of each simulation, we end up with a total of $150$ models after gitsbe finishes its execution.
- **Drabme-specific**:
  - A [`ensemble-wise synergies file`](https://github.com/druglogics/synergy-tutorial/blob/main/ags_cascade_1.0_20211121_150114/ags_cascade_1.0_ensemblewise_synergies.tab), providing the synergy predictions based on gitsbe's models.
  For every drug combination tested (we have $7$ drugs in the [drugpanel file](https://github.com/druglogics/druglogics-synergy/blob/v1.2.1/ags_cascade_1.0/drugpanel), so a total of ${7\choose2} = 21$ drug pairs are tested), drabme calculates a synergy score indicating how synergistic that combination was (more negative means more synergistic), based on the HSA synergy metric.

## How are predictions calculated? {-}

Let's look at an example: drabme [has calculated](https://github.com/druglogics/synergy-tutorial/blob/main/ags_cascade_1.0_20211121_150114/ags_cascade_1.0_ensemblewise_synergies.tab#L11) a synergy score equal to $\approx -0.943$ for the combination `PD-AK` (MEK and AKT inhibitor).

Drabme used gitsbe's $150$ output Boolean models and perturbed them first with the single drug `PD` (changing the equation of the respective target to equal $0$, i.e. `MEK_f = 0`).
Then it calculated all the models' attractors.
The ensemble of models that had at least one attractor were able to provide an [output response](https://druglogics.github.io/druglogics-doc/modeloutputs.html) each, from which drabme calculates the ensemble's average response to the `PD` drug.
Same procedure is followed for the `AK` drug: changing its target's equation appropriately (`AKT_f = 0`) in all $150$ models, calculating their attractors and using only the models that had at least one attractor to calculate the ensemble's average response to drug `PD`.
So we are left with two average single-drug response values, from which we choose as a *reference value* (according to the HSA model we used) **the more negative of the two** (as the values represent cell growth output).

Drabme repeats the previous process by now **changing both equations** in all $150$ models, computing their attractors and finding the ensemble's average growth output response for the `PD-AK` combination.
If this combination value is *lower than the reference value*, the difference is outputted in the ensemble synergies output file.
For our particular case, the difference of the combination's `PK-AD` average response to the minimum of the single-drug average responses in the $150$ model ensemble, is equal to $\approx -0.943$.
Since this value is the lowest in the output [ensemble-wise synergies file](https://github.com/druglogics/synergy-tutorial/blob/main/ags_cascade_1.0_20211121_150114/ags_cascade_1.0_ensemblewise_synergies.tab), `PK-AD` is the most synergistic drug combination out of our simulations.

# Visualize Prediction Performance {-}

In this section we assess the quality of drabme's synergy predictions by plotting ROC and PR curves.

## Observed synergies for CASCADE 1.0 {-}

The synergy prediction problem is a standard binary classification problem (synergy or no synergy, varying a threshold defining how stringent is our definition of synergy).
To properly assess our algorithm's performance we need to compare our results with a true set of drug synergies.
All $21$ drug combinations from the [drugpanel file](https://github.com/druglogics/druglogics-synergy/blob/v1.2.1/ags_cascade_1.0/drugpanel) were experimentally tested in [@Flobak2015], producing a set of $4$ gold standard synergies (see [respective file](https://github.com/druglogics/synergy-tutorial/blob/main/observed_synergies_cascade_1.0)): `PI-PD`, `PI-5Z`, `PD-AK` and `AK-5Z`.

## Workflow {-}

:::{.note}
We will use some `R` code to parse drabme's output prediction scores and visualize our pipeline's synergy prediction performance using the **Receiver Operating Characteristic (ROC)** and the **Precision-Recall (PR)** curves as well as the **Area Under the Curve (AUC)**.
The below code should be run from the root of the [`synergy-tutorial`](https://github.com/druglogics/synergy-tutorial) repository.

We also include a separate [R script](https://github.com/druglogics/synergy-tutorial/blob/main/code.R) which includes all the below code blocks for ease of use.
:::

## Read Output Results {-}

Load the appropriate libraries:
```{r load-lib, class.source = 'fold-show'}
suppressMessages(library(dplyr))
library(tibble)
library(emba)
library(usefun)
library(PRROC)
library(DT)
```

Read the data using the `emba` library [@Zobolas2020]:
```{r read-data, class.source = 'fold-show'}
# Read ensemble-wise synergies file
# `ss` => models trained to steady state
ss_hsa_file = "ags_cascade_1.0_20211121_150114/ags_cascade_1.0_ensemblewise_synergies.tab"
ss_hsa_ensemblewise_synergies = emba::get_synergy_scores(ss_hsa_file)

# Read observed synergies file
observed_synergies_file = 'observed_synergies_cascade_1.0'
observed_synergies = emba::get_observed_synergies(observed_synergies_file)
# 1 (positive/observed synergy) or 0 (negative/not observed) for all tested drug combinations
observed = sapply(ss_hsa_ensemblewise_synergies$perturbation %in% observed_synergies, as.integer)

# Make a data table
pred_hsa = dplyr::bind_cols(ss_hsa_ensemblewise_synergies %>% rename(ss_score = score),
  tibble::as_tibble_col(observed, column_name = "observed"))
```

:::{#ss-pred-tbl}
The prediction results in a table format:
:::
```{r ss-pred-res}
DT::datatable(data = pred_hsa, options = 
  list(pageLength = 7, lengthMenu = c(7, 14, 21), searching = FALSE,
    order = list(list(2, 'asc')))) %>%
  DT::formatRound(columns = 2, digits = 5) %>%
  DT::formatStyle(columns = 'observed',
    backgroundColor = styleEqual(c(0, 1), c('white', '#ffffa1')))
```

## Receiver Operating Characteristic (ROC) {-}

We use the `get_roc_stats()` function from the `usefun` R package [@R-usefun]:
```{r roc-res, fig.align='center', out.width='65%', fig.width=5, fig.height=5, dpi=300, fig.cap='ROC curve (CASCADE 1.0, HSA synergy method, 150 models calibrated to the AGS steady state)', class.source = 'fold-show'}
# Get ROC statistics (`roc_res$AUC` holds the ROC AUC)
roc_res = usefun::get_roc_stats(df = pred_hsa, pred_col = "ss_score", label_col = "observed")

# Plot ROC
my_palette = RColorBrewer::brewer.pal(n = 9, name = "Set1")

plot(x = roc_res$roc_stats$FPR, y = roc_res$roc_stats$TPR,
  type = 'l', lwd = 3, col = my_palette[1], main = 'ROC curve, Ensemble-wise synergies (HSA)',
  xlab = 'False Positive Rate (FPR)', ylab = 'True Positive Rate (TPR)')
legend('bottomright', title = 'AUC', col = my_palette[1], pch = 19,
  legend = paste(round(roc_res$AUC, digits = 2), "Calibrated"), cex = 1.3)
grid(lwd = 0.5)
abline(a = 0, b = 1, col = 'lightgrey', lty = 'dotdash', lwd = 1.2)
```

## Precision-Recall (PR) {-}

We use the `pr.curve()` function from the `PRROC` package [@Grau2015]:
```{r pr-res, fig.align='center', out.width='65%', fig.width=5, fig.height=5, dpi=300, fig.cap='PR curve (CASCADE 1.0, HSA synergy method, 150 models calibrated to the AGS steady state)', class.source = 'fold-show'}
# Get PR statistics (`pr_res$auc.davis.goadrich` holds the PR AUC)
# NOTE: PRROC considers by default that larger prediction values indicate the 
# positive class labeling. For us, the synergy scores belonging to the positive 
# or synergy class (observed = 1) are the lower ones, so we need to
# reverse the scores to correctly calculate the PR curve
pr_res = PRROC::pr.curve(scores.class0 = pred_hsa %>% pull(ss_score) %>% (function(x) {-x}), 
  weights.class0 = pred_hsa %>% pull(observed), curve = TRUE, rand.compute = TRUE)

plot(pr_res, main = 'PR curve, Ensemble-wise synergies (HSA)', 
  auc.main = FALSE, color = my_palette[1], rand.plot = TRUE)
legend('topright', title = 'AUC', col = my_palette[1], pch = 19,
  legend = paste(round(pr_res$auc.davis.goadrich, digits = 2), "Calibrated"))
grid(lwd = 0.5)
```

# Normalization {-}

In this section we show how we can make our predictions better by normalizing them to the predictions of gitsbe models trained to a **random proliferative** [training data profile](https://druglogics.github.io/druglogics-doc/training-data.html#unperturbed-condition---globaloutput-response).
The only change required to run the random model simulation is to change the respective training data file.
We have the required [training data file](https://github.com/druglogics/druglogics-synergy/blob/v1.2.1/ags_cascade_1.0/random_train) already in the `druglogics-synergy` repository so execute the following two commands from its root:

```
cat ags_cascade_1.0/random_train > ags_cascade_1.0/training

java -cp target/synergy-1.2.1-jar-with-dependencies.jar eu.druglogics.synergy.Launcher --inputDir=ags_cascade_1.0
```

:::{.note}
We provide the output simulation results in [this directory](https://github.com/druglogics/synergy-tutorial/tree/main/ags_cascade_1.0_20211128_143028).
:::

Same as before, we read the output data:
```{r read-rnd-data, class.source = 'fold-show'}
# Read ensemble-wise synergies file
# `prolif` => models trained to random proliferative profile
prolif_hsa_file = "ags_cascade_1.0_20211128_143028/ags_cascade_1.0_ensemblewise_synergies.tab"
prolif_hsa_ensemblewise_synergies = emba::get_synergy_scores(prolif_hsa_file)

# check: predictions for the same perturbations
stopifnot(all(prolif_hsa_ensemblewise_synergies$perturbation == ss_hsa_ensemblewise_synergies$perturbation))

# Add random predictions column to the predictions table
pred_hsa = pred_hsa %>%
  tibble::add_column(prolif_score = prolif_hsa_ensemblewise_synergies$score, .before = 'observed')
```

To normalize the steady state calibrated synergy scores, we **subtract the random model predictions** ^[This is normalization in the exponential space, since $\text{norm_score} = x-y$ and $e^{(x-y)} = e^x/e^y$ have the same ranking, $x$ being the vector of calibrated synergy scores ($\text{ss_score}$) and $y$ the random one ($\text{prolif_score}$)]:
```{r norm, class.source = 'fold-show'}
pred_hsa = pred_hsa %>%
  mutate(norm_score = ss_score - prolif_score, .before = 'observed')
```

All the prediction results in a table format:
```{r all-pred-tbl}
DT::datatable(data = pred_hsa, options = 
  list(pageLength = 7, lengthMenu = c(7, 14, 21), searching = FALSE,
    order = list(list(2, 'asc')))) %>%
  DT::formatRound(columns = c(2, 3, 4), digits = 5) %>%
  DT::formatStyle(columns = 'observed',
    backgroundColor = styleEqual(c(0, 1), c('white', '#ffffa1'))) %>%
   DT::formatStyle(columns = 'perturbation', target = 'row',
    backgroundColor = styleEqual(c('BI-PK'), c('#ade2e6')))
```

:::{.green-box}
The **importance of normalization** can be seen by observing how the top 7 predictions by the calibrated models ($\text{ss_score}$) changed in the `norm_score` column.

E.g. `BI-PK` was synergistic according to the steady state calibrated predictions ($\text{ss_score}=-0.56414$) but was **even more synergistic** according to the random model predictions ($\text{prolif_score}=-1.75840$).
Therefore the normalized score **penalizes** such combinations to be more antagonistic ($\text{norm_score = ss_score - prolif_score}=1.19426$), allowing other observed synergies (e.g `AK-5Z`, `PI-5Z`) to 'climb up' to the top predictions list.

Arranging by ascending order the `norm_score` column shows that **the top 5 normalized predictions include all the observed synergies**.
:::

So, as expected, the normalized prediction scores (**Calibrated Normalized**), have better ROC and PR AUCs:
```{r norm-perf-res, fig.show='hold', out.width='50%', fig.width=5, fig.height=5, dpi=300, fig.cap='ROC and PR curve (CASCADE 1.0, HSA synergy method, Calibrated: predictions from 150 models calibrated to the AGS steady state (ss_score), Calibrated Normalized: Calibrated predictions normalized to random models predictions (norm_score))'}
# Get ROC statistics (`roc_res_norm$AUC` holds the ROC AUC)
roc_res_norm = usefun::get_roc_stats(df = pred_hsa, pred_col = "norm_score", label_col = "observed")

# Get PR statistics (`pr_res$auc.davis.goadrich` holds the PR AUC)
pr_res_norm = PRROC::pr.curve(scores.class0 = pred_hsa %>% pull(norm_score) %>% (function(x) {-x}), 
  weights.class0 = pred_hsa %>% pull(observed), curve = TRUE, rand.compute = TRUE)

# ROC figure
plot(x = roc_res_norm$roc_stats$FPR, y = roc_res_norm$roc_stats$TPR,
  type = 'l', lwd = 4, col = my_palette[2], main = 'ROC curve, Ensemble-wise synergies (HSA)',
  xlab = 'False Positive Rate (FPR)', ylab = 'True Positive Rate (TPR)')
lines(x = roc_res$roc_stats$FPR, y = roc_res$roc_stats$TPR, 
  lwd = 3, col = my_palette[1])
legend('bottomright', title = 'AUC', col = my_palette[2:1], pch = 19,
  legend = c(paste(round(roc_res_norm$AUC, digits = 2), "Calibrated Normalized"),
    paste(round(roc_res$AUC, digits = 2), "Calibrated")), cex = 1.3)
grid(lwd = 0.5)
abline(a = 0, b = 1, col = 'lightgrey', lty = 'dotdash', lwd = 1.2)

# PR figure
plot(pr_res_norm, main = 'PR curve, Ensemble-wise synergies (HSA)', 
  auc.main = FALSE, color = my_palette[2], rand.plot = TRUE, lwd = 4)
plot(pr_res, add = TRUE, color = my_palette[1])
legend('bottomleft', title = 'AUC', col = my_palette[2:1], pch = 19,
  legend = c(paste(round(pr_res_norm$auc.davis.goadrich, digits = 2), "Calibrated Normalized"),
    paste(round(pr_res$auc.davis.goadrich, digits = 2), "Calibrated")), cex = 0.8)
grid(lwd = 0.5)
```

# Trying various input configurations {-}

We offer a [script](https://github.com/druglogics/druglogics-synergy/blob/v1.2.1/run_druglogics_synergy.sh) which can be used to run the `druglogics-synergy` module with different input files and configurations.
This script was particularly made to run simulations with the CASCADE 1.0 and 2.0 topologies (using the respective input files), while changing some of gitsbe's and drabme's configuration parameters (e.g. the number of simulations and the synergy calculation method).

In particular, executing the script from the root of the `druglogics-synergy` repository with no changes ^[Note that the CASCADE 2.0 simulations take substantially more time than the CASCADE 1.0 ones, due to a larger topology and drug perturbation set tested.]:

```
./run_druglogics_synergy.sh
```

will run the synergy prediction pipeline (gitsbe and drabme) a total of $16$ times.
Each time, a different combination of configurations will be chosen from:

- $2$ topologies (CASCADE 1.0 or CASCADE 2.0)
- $2$ training data files (train to the curated AGS steady state or a random proliferative profile)
- $2$ different number of genetic algorithm simulations ($50$ or $150$)
- and $2$ synergy calculation methods (HSA or Bliss)

:::{.green-box}
All in all, the given script can be used as a *reference* to extend or build your own and run multiple simulations with our software pipeline based on different input directories and configuration options.
:::

# R session info {-}

```{r session info, comment="", class.source = 'fold-show'}
xfun::session_info()
```

# References {-}
